plantecophys - an R package for analysis and modelling leaf gas exchange data

### 2015-07-06 manuscript.Rmd
New figures, lots of text updated.

---
title: plantecophys - an R package for analysis and modelling of leaf gas exchange data
author: ''
date: ''
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: plos-one.csl
bibliography: references.bib
---

@@

Here I present the R package 'plantecophys', a toolkit to analyse and model leaf gas exchange data, a cornerstone of plant ecophysiological research. Measurements of leaf photosynthesis and transpiration are routinely collected with portable gas exchange instruments, and analyzed with a few key models. These models include the Farquhar-von Caemmerer-Berry model of leaf photosynthesis, the Ball-Berry models of stomatal conductance, and the coupled leaf gas exchange model which combines the supply and demand functions for CO2 in the leaf. The 'plantecophys' R package includes functions both for fitting these models to measurements, as well as simulating from the fitted models to aid in interpreting experimental data. Here I describe the functionality and implementation of the new package, and give examples on its use in plant ecophysiological research. The package makes technically challenging calculations easily accessible for many users. I briefly describe functions for modelling C3 leaf gas exchange with the coupled photosynthesis-stomatal conductance model, fitting of the Farquhar model of photosynthesis to measurements of photosynthesis-CO~2~ response curves ('A-Ci curves'), modelling C4 photosynthesis, numerical solution of optimal stomatal behaviour, and energy balance calculations using the Penman-Monteith equation. The package is freely available on CRAN, and a development version is hosted at http://www.bitbucket.org/remkoduursma/plantecophys.

@@

Wealth of leaf gas exchange data collected in the field of plant ecological physiology. Especially since the advent of portable, easy to use, gas exchange instruments such as the Licor-6400.

These data are used to estimate important plant traits such as Vcmax&Jmax, as well to study responses of stomatal conductance to key environmental drivers.

The photosynthesis model of Farquhar, von Caemmerer and Berry [@farquhar1980] (the 'FvC model') is very widely used in interpreting leaf gas exchange data (cites), up to predicting carbon uptake by vegetation at a global scale. The key prediction of the model is the response of photosynthesis to [CO~2~] in the intercellular airspaces (Ci). In turn, Ci is controlled the result of the balance between diffusion of CO2 through the stomata, and the uptake of CO2 by photosynthesis. To employ the model, it is generally fit to observations of net photosynthesis along a range of [CO2] concentrations.

Various methods exist to fit the model to data, which requires some finesse because the model is not a single non-linear equation, but instead a minimum function that is sometimes difficult to fit (Gu), and sample sizes collected are often small due to time constraints in the field. We do not repeat a description of the Farquhar model here, it has been described many times. But generally it is of the form,

@@

where A~c~ is the gross photosynthesis rate when Rubisco activity is limiting, A~j~ when RuBP-regeneration is limiting, and R~d~ the rate of dark respiration. A~c~ and A~j~ are non-linear functions of the intercellular CO~2~ concentration (C~i~), both of the form $k_1(C_i\Gamma^∗)/(k_2 + C_i)$, where $\Gamma^∗$ is the CO~2~ compensation point without R~d~, and k~1~ and k~2~ are different parameter combinations for A~c~ and A~j~. The details of these functions and the temperature dependence of the parameters are described elsewhere (e.g. Medlyn et al., 2002).
As mentioned above, the parameters Vcmax and Jmax that are embedded in Eq. 1 are not easy to estimate, moreover because the various dependencies of the parameters on leaf temperature must be accounted for.

Once the A-Ci curve is fitted, we can use the Farquhar (FvC) model to simulate the relationship between substrate availability (for which Ci is as a proxy) and rates of photosynthetic CO2 assimilation.

@@

We now have two equations for the photosynthesis rate. The demand-based one (Eq. 1) and the diffusion-based one. At steady state these two should be equal, which can be graphically shown as in Fig. 1 (cf. [@farquhar1982]).

The next step is to include a model of g~s~ that allows for the known response of g~s~ to drivers including vapour pressure deficit and atmospheric CO~2~ concentration (C~a~). The most successful, though empirical, suite of gs models is the Ball-Berry class of models where gs is modelled as a function of A~n~ and D and CO2. This way, effects of leaf temperature and PPFD - both of which are known to affect gs - are modelled through the dependency of photosynthesis on these drivers. Thus the third equation is,

@@

Through Eqs. 1-3 we now have the coupled leaf gas exchange model, which allows modelling of A, g_s and leaf transpiration rate in response to all major environmental drivers (except soil water limitation), and key leaf traits (g1, Vcmax, Jmax, Rd). This coupled leaf gas exchange model can be solved as a quadratic equation, yielding the setpoint C~i~ (Fig. 1), and has found use in many applications (cite). In the next section we describe the various options for f(D) as well as an independent approach assuming that stomatal conductance is optimized to maximize photosynthesis at a given water expenditure (cf. Cowan, Medlyn).

Despite the widespread use of a standard instrument to collect data, tools to analyze the data are scattered and subject to little standardization. Some excel sheets here and there, etc., but we need a solid set of tools in the widely used R language. I here describe the plantecophys package, implemented in the R language.

We have used the code in several of our papers, including [@medlyn2011; @duursma2013; @duursma2014].

@@

The main functions in the package include functions to :

- Fit, summarize, plot and simulate photosynthesis-[CO~2~] response curves (known as 'A-Ci' curves)
- Fit Ball-Berry type models of stomatal conductance
- Estimate optimal stomatal conductance with a numerical implementation of the Cowan-Farquhar hypothesis
- Simulate leaf photosynthesis and transpiration with the coupled leaf gas exchange model
- Estimate leaf temperature from energy balance, when a significant leaf boundary layer is present
- Convert between commonly used units (relative humidity, vapour pressure deficit, dewpoint temperature).

@@

The 'plantecophys' package is implemented in native R, and has no dependencies on other packages. As such it compiles easily, and is highly portable. The source code is maintained with git version control, and is hosted in an online repository (http://www.bitbucket.org/remkoduursma/plantecophys). The repository includes an issue tracker, where users can suggest changes or report bugs.

## Results

@@

We have implemented `fitaci`, which fits the FvC model in two different ways. A widely used published method requires the user to specify the transition of Vcmax to Jmax limitation [@sharkey2007], a process that both prevents batch analysis and is arbitrary. Another method (http://leafweb.ornl.gov/) requires online submission of data and fits the model without much control or knowledge of the fitting process [@gu2010]. Undoubtedly many more implementations of the fitting process have been developed over the years, but few of these are made publicly available (but see e.g. https://github.com/mdekauwe/FitFarquharModel).

Details about it : fits the hyperbolic min function simultaneously. Retrieves estimates of Vcmax, Jmax and Rd (the latter can also be held constant at a known value, in which case only Jmax and Vcmax are fit). Optionally can specify the transition point, so that the Vcmax-limited part is fit independent of the Jmax-limited part. Returns parameter estimates and standard errors, and includes a plot method. Fig 2. is made like this:

```
f <- fitaci(mydata)
plot(f)
```

The function `Aci` simulates the A-Ci curve.

@@

Implementation of several stomatal conductance models, including [@ballberry1987; @leuning1995; @medlyn2011].

@@

The intersection of the supply and demand curves of photosynthesis gives the operating intercellular CO~2~ concentration. This is solved by the `Photosyn` function.

@@

Full numerical optimization of the Cowan-Farquhar hypothesis. We have used this code in [@medlyn2011], and is very similar to that used by [@buckley2014].

@@

We need a publicly available, open source, set of tools to analyze these data which form the cornerstone of plant physiological ecology. Though alternative methods exist to fit A-Ci curves and other details of the implementations can see alternatives, the plantecophys package provides a platform where such methods comparisons can be made. The R language is now widely used, and many users can make changes to the code and recompile the package. 
We have included a set of functions that are key to ecophysiological research. There is certainly scope, and interest, to increase the array of tools included in the package.

@@

# Acknowledgements {.unnumbered}
Thanks to John Drake for extensive testing. David Ellsworth, Belinda Medlyn, Martin de Kauwe are acknowledged for discussion on fitting A-Ci curves. Special thanks to Belinda Medlyn whose code in the MAESTRA model provided inspiration for an early version of the Photosyn function.

@@

```{r fig.width=6, fig.height=6, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure1()
```
** Figure 1**. The intersection of the supply and demand curves of photosynthesis. The `Photosyn` function solves for C~i~ if g~s~, V~cmax~, J~max~ and R~d~ (and other parameters to the FvC model) are known.

```{r fig.width=6, fig.height=6, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure2()
```
**Figure 2**. Standard output from the `fitaci` function. A~net~ is the net photosynthetic rate, C~i~ the intercellular CO~2~ concentration. Symbols are measurements, the line the fitted Farquhar model of photosynthesis.

>>>>

---
title: plantecophys - an R package for analysis and modelling leaf gas exchange data
author: ''
date: ''
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: plos-one.csl
bibliography: references.bib
---

@@

Here I present the R package 'plantecophys', a toolkit to analyse and model leaf gas exchange data, a cornerstone of plant ecophysiological research. Measurements of leaf photosynthesis and transpiration are routinely collected with portable gas exchange instruments, and analyzed with a few key models. These models include the Farquhar-von Caemmerer-Berry model of leaf photosynthesis, the Ball-Berry models of stomatal conductance, and the coupled leaf gas exchange model which combines the supply and demand functions for CO2 in the leaf. The 'plantecophys' R package includes functions both for fitting these models to measurements, as well as simulating from the fitted models to aid in interpreting experimental data. Here I describe the functionality and implementation of the new package, and give examples on its use in plant ecophysiological research.  I briefly describe functions for modelling C3 leaf gas exchange with the coupled photosynthesis-stomatal conductance model, fitting of the Farquhar model of photosynthesis to measurements of photosynthesis-CO~2~ response curves ('A-Ci curves'), modelling C4 photosynthesis, numerical solution of optimal stomatal behaviour, and energy balance calculations using the Penman-Monteith equation. The package makes technically challenging calculations easily accessible for many users and is freely available on CRAN.

@@

Since the advent of portable gas exchange instruments, a wealth of data on leaf gas exchange of CO2 and H~2~O has been collected. These data play a central role in physiological plant ecology, because rates of photosynthesis and transpiration vary widely between plant species, and fluctuate rapidly in response to environmental drivers such as light, humidity and temperature. Not only do leaf gas exchange data allow detailed studies of the underlying plant physiological controls, they are also used to parameterize an important component of land surface models used to predict global water and carbon cycling [e.g. @bonan2014; @dekauwe2015].

\
The photosynthesis model of Farquhar, von Caemmerer and Berry [@farquhar1980] (the 'FvC model') is widely used in interpreting leaf gas exchange data (cites), by providing sensible and comparable metrics of the photosynthetic capacity, and accurately predicting the response of photosynthesis to changes in the CO~2~ concentration inside the leaf air space (C~i~). This widely cited model is embedded in many global vegetation models (cite). The key prediction of the model is the response of photosynthesis to [CO~2~] in the intercellular airspaces (Ci). To employ the model, it is generally fit to observations of net photosynthesis along a range of [CO2] concentrations, yielding well-known measures of photosynthetic capacity (V~cmax~ and J~max~).

\
We do not repeat a detailed description of the FvC model here, as it has been described many times. But generally it is of the form,

@@

\
where A~c~ is the gross photosynthesis rate when Rubisco activity is limiting, A~j~ when RuBP-regeneration is limiting, and R~d~ the rate of dark respiration. A~c~ and A~j~ are non-linear functions of the intercellular CO~2~ concentration (C~i~), both of the form $k_1(C_i\Gamma^∗)/(k_2 + C_i)$, where $\Gamma^∗$ is the CO~2~ compensation point without R~d~, and k~1~ and k~2~ are different parameter combinations for A~c~ and A~j~. The details of these functions and the temperature dependence of the parameters are described elsewhere [@medlyn2002]. The use of C~i~ in these equations is routine, but actually it should be Cc, considering that CO2 in the chloroplast is not equal to that in the intercellular air space.

\
Fitting the model to data requires some finesse because the model is not a single non-linear equation, but instead a minimum function of two non-linear equations that is sometimes difficult to fit. Moreover, sample sizes collected are often small due to time constraints in the field. A widely used published method requires the user to specify the transition of Vcmax to Jmax limitation [@sharkey2007], a process that both prevents batch analysis and is arbitrary. Another method (http://leafweb.ornl.gov/) requires online submission of data and fits the model without much control or knowledge of the fitting process [@gu2010]. Undoubtedly many more implementations of the fitting process have been developed over the years, but few of these are made publicly available (but see available online tools [@dekauwefitfarquhar; @landfluxtools]).

\
Through Eq. 1, we have a dependency of photosynthesis on the availability of the substrate, Ci. To estimate Ci itself, we must have an estimate of the stomatal conductance to CO~2~. From Fick's law, we can relate A~n~ to g~s~ and C~i~ as,

@@

\
where g~s~ is the conductance to CO~2~. We now have two equations for the photosynthesis rate. The 'demand function' is given by Eq. 1, and the 'supply function' in Eq. 2. At steady state these two equation should be equal, which can be graphically shown as in Fig. 1 (cf. [@farquhar1982]).

\
Because g~s~ itself responds to environmental drivers, we must then estimate this dependence with another model. The most successful, though empirical, g~s~ models is the Ball-Berry [@ballberry1987] class of models. This model posits an entirely empirical equation that describes the response of g~s~ to air humidity, CO~2~ and the rate of photosynthesis. This way, effects of leaf temperature and PPFD - both of which are known to affect g~s~ - are modelled through the dependency of A~n~ on these drivers. Thus the third equation is,

@@

\
where D is the vapour pressure deficit (kPa), g~0~ and g~1~ are empirical parameters. An alternative approach to modelling g~s~ is through the hypothesis that stomata act optimally in the sense that they maximize photosynthesis while minimizing water loss. This hypothesis was first developed by Cowan and Farquhar [@cowan1977] and has seen many applications. Medlyn et al. [@medlyn2011] showed that the optimality hypothesis, when coupled to the FvC model, leads to an expression analogous to the Ball-Berry type models (Eq. 3), but with a different D response function (f(D) in Eq. 3) than the original Ball-Berry model, or yet another version introduced by Leuning [@leuning1995].

\
Finally we can combine the biochemical demand function of photosynthesis (Eq. 1) with the supply function (Eq. 2) and an expression for the dependency of g~s~ on environmental drivers (Eq. 3). This 'coupled' leaf gas exchange model [@leuning1990; @collatz1991] is implemented in many process-based ecosystem models [@wangl1998; @bonan2014].
This model allows prediction of A, g_s and leaf transpiration rate in response to all major environmental drivers (except soil water limitation), and key leaf traits (g1, Vcmax, Jmax, Rd).

\
Despite the widespread use of the FvC model and the coupled leaf gas exchange model, tools to analyze the data are scattered and subject to little standardization. Fitting the FvC model to CO~2~ response curves is a standard procedure but different methods can yield different parameter values, making comparisons difficult. The coupled leaf gas exchange model is not straightforward to implement, and I do not know of an open-source implementation that is publicly available. I here describe the plantecophys package, implemented in the R language [@rfoundation]. The code is freely available (without restrictions), and managed with a version control system. I have developed the code base in my work on leaf and canopy modelling of photosynthesis and stomatal conductance, with many additions based on user requests [@medlyn2011; @barton2012; @medlyn2013; @duursma2013; @duursma2014].

@@

The main functions in the package are summarized in Table 1.

```{r results='asis', echo=FALSE, message=FALSE, warn=FALSE}
f <- c("`fitaci`","`fitBB`","`FARAO`","`Photosyn`","`PhotosynEB`","`RHtoVPD` etc.")
d <- c("Fit, summarize, plot and simulate photosynthesis-[CO~2~] response curves (A-Ci curves)",
"Fit Ball-Berry type models of stomatal conductance",
"Estimate optimal stomatal conductance with a numerical implementation of the Cowan-Farquhar hypothesis",
"Simulate leaf photosynthesis and transpiration with the coupled leaf gas exchange model",
"Estimate leaf temperature from energy balance, when a significant leaf boundary layer is present",
"Convert between commonly used units (relative humidity, vapour pressure deficit, dewpoint temperature)")
df <- data.frame(Function=f, Description=d)

# library(knitr)
# kable(df, format="markdown",
#       caption="Table 1. Main functions in the plantecophys package. Each function has its own detailed help page  including example code.")

library(ascii)
suppressWarnings(print(ascii(df, caption="Table 1. Main functions in the plantecophys package.",
            include.rownames=FALSE),
      type='pandoc'))
```

@@

The 'plantecophys' package is implemented in native R, and has no dependencies on other packages. As such it builds easily, and is highly portable. The source code is maintained with git version control, and is hosted in an online repository (http://www.bitbucket.org/remkoduursma/plantecophys). The repository includes an issue tracker, where users can suggest changes or report bugs.

## Results and Discussion

@@

The `fitaci` function fits the FvC model, yielding estimates of V~cmax~, J~max~ and R~d~ and their standard errors. Optionally R~d~ can be provided as a known value, otherwise it is estimated from the A-Ci curve. The user does not have to provide the transition point (see Fig. 1), as this is estimated by `fitaci` automatically. It is however an option to fix the transition point, which may be helpful to check whether the best fit was achieved. Standard errors are estimated with the usual methods for non-linear regression (`nls` function in base R). Instead of fitting the minimum function (Eq. 1) it is actually far more practical to use a hyperbolic minimum, which avoids a discontinuity at the transition from A~c~ to A~j~ limitation.

\
The function takes a dataframe as input, which includes A~n~, C~i~ and optionally leaf temperature (T~leaf~) and photosynthetically active radiation (PAR). It then returns a special object which has a print and plot method.

```
# Fit FvC model
f <- fitaci(mydata)

# Make standard plot
plot(f)
```

\
The output of the above example is shown in Fig. 2. Additionally, `fitacis` can be sued to fit many curves at once, for example one for each species or site in a dataset. The function `Aci` simulates the A-Ci curve, that is, it returns A~n~ when C~i~ is known.

@@

The straightforward `fitBB` function provides an interface to non-linear or linear regression to fit one of three stomatal conductance models [@ballberry1987; @leuning1995; @medlyn2011]. This yields estimates of g~1~ and (optionally ) g~0~, which are necessary inputs to the coupled leaf gas exchange model.

@@

The intersection of the supply and demand curves of photosynthesis gives the operating intercellular CO~2~ concentration. This is solved by the `Photosyn` function. This flexible interface can be used to either a) estimate A~n~ when C~i~ is known (`Photosyn(Ci=...)`, equivalent to `Aci(...)`), b) estimate A~n~ when g~s~ is known (`Photosyn(GS=...)`) (cf. Fig. 2) or c) solve C~i~ from the coupled leaf gas exchange model (Eqs. 1-3).

- example response curves
- energy balance (cite [@wang1998; @buckley2014]) - brief on details.

@@

Full numerical optimization of the Cowan-Farquhar hypothesis. I have used this code in [@medlyn2011], and it is very similar to that presented (but not made publicly available) by [@buckley2014]. It is visualized in Figure 4.

@@

We need a publicly available, open source, set of tools to analyze leaf gas exchange data, as these data form a cornerstone of plant physiological ecology. Though alternative methods exist to fit A-C~i~ curves, the plantecophys package provides a platform where such methods comparisons can be made. As such I do not claim that `fitaci` is the only or best method available. The coupled leaf gas exchange model is widely used but no implementations are publicly available. The plantecophys package fills this void.

The R language is now widely used, and more advanced users can make changes to the code and rebuild the package. Through the git version control system it is straightforward to include new changes into the central repository.

@@

# Acknowledgements {.unnumbered}
David Ellsworth, Belinda Medlyn, Martin de Kauwe are acknowledged for discussion on fitting A-Ci curves. Special thanks to Belinda Medlyn whose code in the MAESTRA model provided inspiration for an early version of the Photosyn function. Thanks to John Drake for testing.

@@

```{r fig.width=6, fig.height=5, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure1()
```

** Figure 1**. Leaf photosynthesis - CO~2~ response curve as modelled with the FvC model.

```{r fig.width=6, fig.height=5, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
suppressWarnings(figure2())
```

** Figure 2**. The intersection of the supply and demand curves of photosynthesis. The `Photosyn` function solves for C~i~ if g~s~, V~cmax~, J~max~ and R~d~ (and other parameters to the FvC model) are known.

```{r fig.width=6, fig.height=5, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure3()
```

**Figure 3**. Standard output from the `fitaci` function. A~net~ is the net photosynthetic rate, C~i~ the intercellular CO~2~ concentration. Symbols are measurements, the line the fitted Farquhar model of photosynthesis.

```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure4()
```

**Figure 4**. Visualization of the optimal model of stomatal conductance. Provided we have an estimate of the 'cost of water' ($\lambda$, mol C mol H~2~O^-1^), stomata act to maximize photosynthesis minus transpiration. In (a), individual curves at a range of values for the vapour pressure deficit (D) plots $A-\lambda*E$ versus the C~i~, demonstrating that an optimum exists. The ~FARAO~ function finds this optimum numerically and calculates A~n~ and g~s~. The corresponding response of g~s~ to D is shown in panel (b).
